---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# RMSTdesign

<!-- badges: start -->
<!-- badges: end -->

The goal of `RMSTdesign` is to make it easy to design clinical trials with the difference in restricted mean survival time as the primary endpoint. This includes calcuating asymptotic power or sample size under user-specified design parameters and performing simulations to determine the operating characteristics of the test of restricted mean survival time.  

## Installation

You can install `RMSTdesign` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("anneae/RMSTdesign")
```

## Documentation
Documentation and vignettes are available [here](https://anneae.github.io/RMSTdesign/).

## Example

In this basic example, we specify that the control group is expected to have a median survival of 3 years, the treatment group is expected to have a median survival of 4 years, and we plan to accrue patients over one year and follow them until the last patient has 3 years of followup. Also, we select $\tau$ = 3 years as a clinically meaningful restriction time. The code below determines the required sample size for a trial with 80% power. 

```{r example}
library(RMSTdesign)
con<-survdef(times = 3, surv = 0.5) 
trt<-survdef(times = 4, surv = 0.5)

RMSTpow(con, trt, k1 = 1, k2 = 3, tau = 3, power = 0.8)

```
