---
title: "RMSTdesign-package-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RMSTdesign-package-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(RMSTdesign)
library(SSRMST)
library(knitr)
library(PWEALL)
library(survival)
```

This vignette validates the internal functions of the `RMSTdesign` package by comparing the results to those obtained by plugging design parameters directly into simple formulas. This vignette also confirms that the asymptotic power and pKME returned by the `RMSTpow` function are close to their empirical conterparts across several scenarios. Along with the results in our paper[^1] (Tables 1 and 3 which show that the power of the RMST-based test and pKME calculated using `RMSTpow` match their empirical counterparts across 4 scenarios, various values of k1, k2, tau and n, and a table in Appendix I which shows that the empirical variance of the RMST estimator matches the variance calculated using the internal function `evar` from `RMSTdesign`), the vignette "Compare-to-SSRMST" (which shows that results from `RMSTdesign` match simulation results from the `SSRMST` package[^2] in 4 scenarios), and the vignette "GOG111-Trial", which shows that `RMSTpow` returns almost the same sample size as the one calculated by Royston and Parmar[^3] (with the difference likely due to different approaches for calculating the restricted mean survival difference and its variance), these results validate that the calculations in the `RMSTdesign` package are correct and that the package can be used for trial design.  

##Approach to validating internal functions
The internal functions implement equations 1, 3 and 4 from our paper[^1]. We can show that in simple settings, these equations simplify to simple formulas in terms of the design parameters.Below, we have presented the simple formulas, and then shown that the results of the internal functions match the results obtained by plugging in the design parameters, for several combinations of the design parameters. 

##Validating the evar function 
The `evar` function calculates the variance of the estimated difference in RMST according to equation 1 in our paper. We will confirm that it gives the correct result in a simple setting. In the setting of an exponential survival distribution and no censoring before time $\tau$, the asymptotic variance of the estimated mean survival time is equal to 

$$\frac{1}{n\lambda^2}(1-\exp(-\lambda\tau)\times\{\exp(-\lambda\tau)+2\lambda\tau\})$$
 
```{r, echo=F}
evar<-function(survdefo,  k1, k2, tau, ttime, n){
    P_enroll<-function(t,x) {
        if (t<=x) 0
        else if (x<t & t<=(k1+x))  (t-x)/k1
        else if (k1+x <t & t<= k1+k2) 1
        else if (k1+x <t & all.equal(t, k1+k2)) 1}
    toint<-function(u) sapply(u, function(x) integrate(survdefo$S, lower = x, upper = tau)$value^2*survdefo$h(x)/(n*survdefo$S(x)*P_enroll(ttime, x)))
    integrate(toint, lower = 0, upper = tau)$value
}
```

Below, the variance at the end of the trial (calculated with `evar`) is compared to the variance calculcated according to the formula above for several combinations of design parameters:
```{r}
n<-200; lambda<-0.07; tau <-10
evar(survdef(haz = lambda), 0, tau, tau, tau, n)
(1/(n*lambda^2))*(1-exp(-lambda*tau)*(exp(-lambda*tau)+2*lambda*tau))

n<-200; lambda<-0.035; tau <-10
evar(survdef(haz = lambda), 0, tau, tau, tau, n)
(1/(n*lambda^2))*(1-exp(-lambda*tau)*(exp(-lambda*tau)+2*lambda*tau))

n<-200; lambda<-0.07; tau <-20
evar(survdef(haz = lambda), 0, tau, tau, tau, n)
(1/(n*lambda^2))*(1-exp(-lambda*tau)*(exp(-lambda*tau)+2*lambda*tau))

n<-400; lambda<-0.07; tau <-10
evar(survdef(haz = lambda), 0, tau, tau, tau, n)
(1/(n*lambda^2))*(1-exp(-lambda*tau)*(exp(-lambda*tau)+2*lambda*tau))
```

##Validating the powfn function
The `powfn` function calculates the power of the one-sided test of the difference in RMST according to equation 3 in our paper. We will confirm that it gives the correct result in a simple setting. In the setting of exponential survival distributions in each group and no censoring before time $\tau$, the sampling distribution of the test statistic Z is has a mean of 
$$\mu=\frac{\frac{1-p^r}{r}-(1-p)}{\frac{1}{\sqrt(n/2)}\sqrt(\frac{1}{r^2}(1-p^r\{p^r+2r(-\log(p))\})+1-p\{p+2(-\log(p))\})}$$

where $r$ is the pazard ratio and $p$ is the survival probability at time $\tau$ in the control group.

```{r, echo = F}
powfn<-function(survdefC, survdefT, k1, k2, tau, n, alpha){
    RMST_truediff<-integrate(function(x) survdefT$S(x)-survdefC$S(x),
                             lower = 0, upper = tau)$value
    RMST_trueSE<-sqrt(evar(survdefT, k1, k2, tau, k1+k2, n/2)+
                          evar(survdefC, k1, k2, tau, k1+k2, n/2))
    return(1-pnorm(qnorm(1-alpha)-RMST_truediff/RMST_trueSE))
}
```

Below, the power at the end of the trial (calculated with `powfn`) is compared to the power calculcated according to the formula above for several combinations of design parameters:
```{r}
n<-200; lambda<-0.07; r<-0.5; tau <-10; alpha = 0.025
p<-survdef(haz = lambda)$S(tau)
mu_num<-(1-p^r)/r-(1-p)
mu_denom<-(n/2)^(-1/2)*sqrt(r^-2*(1-p^r*(p^r+2*r*-log(p)))+(1-p*(p+2*-log(p))))
1-pnorm(qnorm(1-alpha)-mu_num/mu_denom)
powfn(survdef(haz = lambda), survdef(haz = r*lambda), k1=0, k2=tau, 
      tau = tau, n = n, alpha=alpha)

n<-100; lambda<-0.07; r<-0.5; tau <-10; alpha = 0.025
p<-survdef(haz = lambda)$S(tau)
mu_num<-(1-p^r)/r-(1-p)
mu_denom<-(n/2)^(-1/2)*sqrt(r^-2*(1-p^r*(p^r+2*r*-log(p)))+(1-p*(p+2*-log(p))))
1-pnorm(qnorm(1-alpha)-mu_num/mu_denom)
powfn(survdef(haz = lambda), survdef(haz = r*lambda), k1=0, k2=tau, 
      tau = tau, n = n, alpha=alpha)

n<-200; lambda<-0.07; r<-0.9; tau <-10; alpha = 0.025
p<-survdef(haz = lambda)$S(tau)
mu_num<-(1-p^r)/r-(1-p)
mu_denom<-(n/2)^(-1/2)*sqrt(r^-2*(1-p^r*(p^r+2*r*-log(p)))+(1-p*(p+2*-log(p))))
1-pnorm(qnorm(1-alpha)-mu_num/mu_denom)
powfn(survdef(haz = lambda), survdef(haz = r*lambda), k1=0, k2=tau, 
      tau = tau, n = n, alpha=alpha)

n<-200; lambda<-0.035; r<-0.5; tau <-10; alpha = 0.025
p<-survdef(haz = lambda)$S(tau)
mu_num<-(1-p^r)/r-(1-p)
mu_denom<-(n/2)^(-1/2)*sqrt(r^-2*(1-p^r*(p^r+2*r*-log(p)))+(1-p*(p+2*-log(p))))
1-pnorm(qnorm(1-alpha)-mu_num/mu_denom)
powfn(survdef(haz = lambda), survdef(haz = r*lambda), k1=0, k2=tau, 
      tau = tau, n = n, alpha=alpha)

n<-200; lambda<-0.07; r<-0.5; tau <-10; alpha = 0.05
p<-survdef(haz = lambda)$S(tau)
mu_num<-(1-p^r)/r-(1-p)
mu_denom<-(n/2)^(-1/2)*sqrt(r^-2*(1-p^r*(p^r+2*r*-log(p)))+(1-p*(p+2*-log(p))))
1-pnorm(qnorm(1-alpha)-mu_num/mu_denom)
powfn(survdef(haz = lambda), survdef(haz = r*lambda), k1=0, k2=tau, 
      tau = tau, n = n, alpha=alpha)

```

##Validating the RMSTeval function
The `RMSTeval` function calculates the probability the difference in the restricted mean survival time at time $\tau$ will be estimable using the Kaplan-Meier estimator, equation 4 in our paper. Consider a very simple situation where survival in each group is equal to 1 until time $\tau$ and $\tau<k_1$. Then we get the probability $p_{KME}$ is estimable in each group is equal to the following, where $g(x)$ is the hazard function for the censoring distribution and $T$ and $C$ are the random death and censoring times: 
$$p_{KME}=1-n\int_0^\tau \{P(\min(T,C)<x)\}^{n-1}g(x)S(x)dx$$
$$= 1-n\int_0^\tau \left(\frac{x}{k_1}\right)^{n-1}\frac{1}{k_1}dx=1-\left(\frac{\tau}{k_1}\right)^n$$
If we let $n$ represent the total sample size, then the sample size in each group is $n/2$ The probability that we can estimate the RMST in both groups with the Kaplan-Meier estimator is the product of the probability we can estimate it in each group:
$$\left[1-\left(\frac{\tau}{k_1}\right)^{n/2}\right]^2$$
```{r, echo=F}
RMSTeval<-function(survdefC, survdefT, k1, k2, tau, n){
    if (tau<=k2) return(1)
    P_x_followup<-function(x){ # density for exactly x amount of followup
        if (x<k2) 0
        else if (x<=(k1+k2)) 1/(k1)
        else 0
    }
    CDF_x_followup<-function(x){ # CDF for x amount of followup
        if (x<k2) 0
        else if (x<=(k1+k2)) (x-k2)/(k1)
        else 1
    }
    # control group
    toint<-function(u) sapply(u, function(t) (1-(1-CDF_x_followup(t))*survdefC$S(t))^(n/2-1)*P_x_followup(t)*survdefC$S(t))
    peval_c<-1-(n/2)*integrate(toint, lower = k2, upper = min(k1+k2, tau))$value
    # trt group
    toint<-function(u) sapply(u, function(t) (1-(1-CDF_x_followup(t))*survdefT$S(t))^(n/2-1)*P_x_followup(t)*survdefT$S(t))
    peval_t<-1-(n/2)*integrate(toint, lower = k2, upper = min(k1+k2, tau))$value
    return(peval_c*peval_t)
}
```
Below, the probability we can use the Kaplan-Meier estimator to estimate the RMST difference at the end of the trial (calculated with `RMSTeval`) is compared to the probability calculcated according to the formula above for several combinations of design parameters:
```{r}
k1=10;tau = 9.9; n=500
RMSTeval(survdef(haz = 0), survdef(haz = 0),k1 = k1, k2 = 0, tau = tau, n=n)
(1-(tau/k1)^(n/2))^2

k1=10;tau = 9.9; n=200
RMSTeval(survdef(haz = 0), survdef(haz = 0),k1 = k1, k2 = 0, tau = tau, n=n)
(1-(tau/k1)^(n/2))^2

k1=12; tau = 9.9; n=500
RMSTeval(survdef(haz = 0), survdef(haz = 0),k1 = k1, k2 = 0, tau = tau, n=n)
(1-(tau/k1)^(n/2))^2

k1=10; tau = 9.8; n=500
RMSTeval(survdef(haz = 0), survdef(haz = 0),k1 = k1, k2 = 0, tau = tau, n=n)
(1-(tau/k1)^(n/2))^2

k1=10; tau = 10; n=500
RMSTeval(survdef(haz = 0), survdef(haz = 0),k1 = k1, k2 = 0, tau = tau, n=n)
(1-(tau/k1)^(n/2))^2

```

##Simulation notes
The `RMSTpow` function can do simulations and report empirical power and $p_{KME}$ using th `sim=T` option. However, we thought that it would be better to validate the asymptotic power and $p_{KME}$ calculated using `RMSTpow` by comparing them to empirical quantities calcualted independent of the `RMSTpow` function. The simulation code used to calculate the empirical quantities is shown below. We use `rpwe` from the `PWEALL` package and `rweibull` to generate the simulated data for piecewise expoenential and Weibull distributions, respectively. We use`summary.survfit` from the `survival` packages with the `rmean` option to calculate the RMST difference estimate and the variance estimate. Empirical quantities will be calculated using 1000 simulations. 

##Simulations for scenario 4, but varying the hazard rate and hazard ratio 
The plots in Figure S9, Scenario 4 show the asymptotic power of the RMST-based test across a range of control group hazard rates and hazard ratios, including the specific design parameter combinations used below. In the original Scenario 4, the hazard rate was about 0.07 and the hazard ratio was 0.67 for the first 15 months, then 1.2. Below, we will consider 4 values of the hazard rate and 4 values of the hazard ratio over the first 15 months. The hazard ratio after 15 months will remain 1.2. 
```{r}
hrate<-c(.05, .1, .15, .2, .07, .07, .07, .07)
hratio<-c(.67, .67, .67, .67, .25, .5, .75, 1)
k1<-24
k2<-36
tau<-48
n<-780
alpha = 0.05
M<-1000
restable<-as.data.frame(matrix(ncol=8, nrow = 8))
colnames(restable)<-c('Haz rate','HR',
                      'Asympt RMST power sup','Asympt RMST power inf','Asymp pKME',
                      'Emp RMST power sup','Emp RMST power inf','Emp pKME')
restable[,1]<-hrate
restable[,2]<-hratio

for (i in 1:8){
    scen1<-survdef(haz = hrate[i])
    scen4T<-survdef(haz = c(hratio[i], 1.2)*hrate[i], times = 15)
    res<-RMSTpow(scen1, scen4T, k1, k2, tau, n = n, two.sided = T)
    restable[i,3]<-res$powerRMSTToverC
    restable[i,4]<-res$powerRMSTCoverT
    restable[i,5]<-res$pKME
    simres<-matrix(nrow = M, ncol = 3)
    for (m in 1:M){
        grp<-c(rep('c', n/2), rep('t', n/2))
        enrol<-runif(n, min=0, max = k1)
        con<-rexp(n/2, rate = hrate[i])
        trt<-rpwe(n/2, rate = c(hratio[i]*hrate[i], 1.2*hrate[i]), tchange = c(0,15))$r
        event<-c(con, trt)+enrol< k1+k2
        time<-pmin(c(con, trt),  k1+k2-enrol)
        surv_res<-survfit(Surv(time, event)~grp, timefix = F)
        c.time<-surv_res$time[1:surv_res$strata[1]]
        c.cens<-surv_res$n.censor[1:surv_res$strata[1]]
        t.time<-surv_res$time[surv_res$strata[1]+1:surv_res$strata[2]]
        t.cens<-surv_res$n.censor[surv_res$strata[1]+1:surv_res$strata[2]]
        
        simres[m,3]<-T # is RMST difference estimable at time tau?
        if (c.cens[length(c.cens)]>0 & tau > max(c.time)) simres[m,3]<-F
        if (t.cens[length(c.cens)]>0 & tau > max(t.time)) simres[m,3]<-F
        summ<-summary(surv_res, rmean = tau)$table
        RMST_diff<-summ[2,5]-summ[1,5]
        simres[m,1]<-RMST_diff/(sqrt(summ[2,6]^2 + summ[1,6]^2))>qnorm(1-alpha/2)
        simres[m,2]<-(RMST_diff)/(sqrt(summ[2,6]^2 + summ[1,6]^2))<qnorm(alpha/2)
    }
    restable[i,6]<-mean(simres[,1])
    restable[i,7]<-mean(simres[,2])
    restable[i,8]<-mean(simres[,3])
}
```

In the table below, Asmypt indicates asymptotic (calculated using our package) and Emp indicates empirical (calculated based on 1000 simulations). Sup indicates the power to conclude treatment is superior to control and Inf indicates the power to conclude that control is superior to treatment. We see that the asymptotic quantities are close to their asymptotic counterparts. 
```{r, echo=F}
kable(restable)
```

##Simulation for crossing survival curves, treatment group is favored initially
The scenario here is as in Figure S6 with the curves crossing at 20 months. We will consider various values of k1, k2 and tau
```{r}
scenC<-survdef(surv = .5, times = 10)
# Want to find b such that scen1C$S(t) = exp(-(t/b)^a)
# This will give the webull surve with shape = a that crosses scen1C at time = t
a<- 1.5# shape param
t<-20
b<- t/(-log(scenC$S(t)))^(1/a)
scenT<-survdefWeibull(shape = a, scale = b)
plotsurvdef(scenC, scenT, xupper = 60)
```

```{r}
k1vec<-c(12,36,24,24,24,24)
k2vec<-c(36,36,12,24,36,36)
tauvec<-c(36,36,36,36,24,48)

n<-780
alpha = 0.05
M<-1000
restable<-as.data.frame(matrix(ncol=9, nrow = 6))
colnames(restable)<-c('k1','k2','tau',
                      'Asympt RMST power sup','Asympt RMST power inf','Asymp pKME',
                      'Emp RMST power sup','Emp RMST power inf','Emp pKME')
restable[,1]<-k1vec
restable[,2]<-k2vec
restable[,3]<-tauvec

for (i in 1:6){
    res<-RMSTpow(scenC, scenT, k1vec[i], k2vec[i], tauvec[i], n = n, two.sided = T)
    restable[i,4]<-res$powerRMSTToverC
    restable[i,5]<-res$powerRMSTCoverT
    restable[i,6]<-res$pKME
    simres<-matrix(nrow = M, ncol = 3)
    for (m in 1:M){
        grp<-c(rep('c', n/2), rep('t', n/2))
        enrol<-runif(n, min=0, max = k1vec[i])
        con<-rexp(n/2, rate = -log(.5)/10)
        trt<-rweibull(n/2, shape=a, scale = b)
        event<-c(con, trt)+enrol< k1vec[i]+k2vec[i]
        time<-pmin(c(con, trt),  k1vec[i]+k2vec[i]-enrol)
        surv_res<-survfit(Surv(time, event)~grp, timefix = F)
        c.time<-surv_res$time[1:surv_res$strata[1]]
        c.cens<-surv_res$n.censor[1:surv_res$strata[1]]
        t.time<-surv_res$time[surv_res$strata[1]+1:surv_res$strata[2]]
        t.cens<-surv_res$n.censor[surv_res$strata[1]+1:surv_res$strata[2]]
        
        simres[m,3]<-T # is RMST difference estimable at time tau?
        if (c.cens[length(c.cens)]>0 & tauvec[i] > max(c.time)) simres[m,3]<-F
        if (t.cens[length(c.cens)]>0 & tauvec[i] > max(t.time)) simres[m,3]<-F
        summ<-summary(surv_res, rmean = tauvec[i])$table
        RMST_diff<-summ[2,5]-summ[1,5]
        simres[m,1]<-RMST_diff/(sqrt(summ[2,6]^2 + summ[1,6]^2))>qnorm(1-alpha/2)
        simres[m,2]<-(RMST_diff)/(sqrt(summ[2,6]^2 + summ[1,6]^2))<qnorm(alpha/2)
    }
    restable[i,7]<-mean(simres[,1])
    restable[i,8]<-mean(simres[,2])
    restable[i,9]<-mean(simres[,3])
}

```

In the table below, Asmypt indicates asymptotic (calculated using our package) and Emp indicates empirical (calculated based on 1000 simulations). Sup indicates the power to conclude treatment is superior to control and Inf indicates the power to conclude that control is superior to treatment. We see that the asymptotic quantities are close to their asymptotic counterparts. 
```{r, echo=F}
kable(restable)
```

##Simulation for crossing survival curves, control group is favored intially
The scenario here is as in Figure S7 with the curves crossing at 20 months. We will consider various values of k1, k2 and tau
```{r}
scenC<-survdef(surv = .5, times = 10)
# Want to find b such that scen1C$S(t) = exp(-(t/b)^a)
# This will give the webull surve with shape = a that crosses scen1C at time = t
a<- .67# shape param
t<-20
b<- t/(-log(scenC$S(t)))^(1/a)
scenT<-survdefWeibull(shape = a, scale = b)
plotsurvdef(scenC, scenT, xupper = 60)
```

```{r}
k1vec<-c(12,36,24,24,24,24)
k2vec<-c(36,36,12,24,36,36)
tauvec<-c(36,36,36,36,24,48)

n<-780
alpha = 0.05
M<-1000
restable<-as.data.frame(matrix(ncol=9, nrow = 6))
colnames(restable)<-c('k1','k2','tau',
                      'Asympt RMST power sup','Asympt RMST power inf','Asymp pKME',
                      'Emp RMST power sup','Emp RMST power inf','Emp pKME')
restable[,1]<-k1vec
restable[,2]<-k2vec
restable[,3]<-tauvec

for (i in 1:6){
    res<-RMSTpow(scenC, scenT, k1vec[i], k2vec[i], tauvec[i], n = n, two.sided = T)
    restable[i,4]<-res$powerRMSTToverC
    restable[i,5]<-res$powerRMSTCoverT
    restable[i,6]<-res$pKME
    simres<-matrix(nrow = M, ncol = 3)
    for (m in 1:M){
        grp<-c(rep('c', n/2), rep('t', n/2))
        enrol<-runif(n, min=0, max = k1vec[i])
        con<-rexp(n/2, rate = -log(.5)/10)
        trt<-rweibull(n/2, shape=a, scale = b)
        event<-c(con, trt)+enrol< k1vec[i]+k2vec[i]
        time<-pmin(c(con, trt),  k1vec[i]+k2vec[i]-enrol)
        surv_res<-survfit(Surv(time, event)~grp, timefix = F)
        c.time<-surv_res$time[1:surv_res$strata[1]]
        c.cens<-surv_res$n.censor[1:surv_res$strata[1]]
        t.time<-surv_res$time[surv_res$strata[1]+1:surv_res$strata[2]]
        t.cens<-surv_res$n.censor[surv_res$strata[1]+1:surv_res$strata[2]]
        
        simres[m,3]<-T # is RMST difference estimable at time tau?
        if (c.cens[length(c.cens)]>0 & tauvec[i] > max(c.time)) simres[m,3]<-F
        if (t.cens[length(c.cens)]>0 & tauvec[i] > max(t.time)) simres[m,3]<-F
        summ<-summary(surv_res, rmean = tauvec[i])$table
        RMST_diff<-summ[2,5]-summ[1,5]
        simres[m,1]<-RMST_diff/(sqrt(summ[2,6]^2 + summ[1,6]^2))>qnorm(1-alpha/2)
        simres[m,2]<-(RMST_diff)/(sqrt(summ[2,6]^2 + summ[1,6]^2))<qnorm(alpha/2)
    }
    restable[i,7]<-mean(simres[,1])
    restable[i,8]<-mean(simres[,2])
    restable[i,9]<-mean(simres[,3])
}

```

In the table below, Asmypt indicates asymptotic (calculated using our package) and Emp indicates empirical (calculated based on 1000 simulations). Sup indicates the power to conclude treatment is superior to control and Inf indicates the power to conclude that control is superior to treatment. We see that the asymptotic quantities are close to their asymptotic counterparts. 
```{r, echo=F}
kable(restable)
```

[^1]: Eaton A, Therneau T and Le-Rademacher J. Designing Clinical Trials with (Restricted) Mean Survival Time Endpoint – Practical Considerations.
[^2]: https://cran.r-project.org/web/packages/SSRMST/vignettes/vignette-ssrmst.html
[^3]: Royston P and Parmar MK. Restricted mean survival time: an alternative to the hazard ratio for the design and analysis of randomized trials with a time-to-event outcome. BMC Med Res Methodol 2013; 13: 152. 
